introduction

### 2019-03-18 introduction.Rmd
proof intro 1

Australia has changed in many ways over the last two decades. Rising house prices, country-wide improvements in education, an ageing population, and a decline in religious affiliation, are just a few facets of the country's evolving socio-demographic characteristics[^1]. At the same time, political power has moved back and forth between the two major parties. In the 2007 and 2010 federal elections, the Australian Labor Party (Labor) was victorious, whereas the 2001, 2004, 2013 and 2016 elections were won by the Liberal National coalition (Liberal). The two-party preferred vote, a measure of support between these two parties, fluctuated between 47.3% and 53.5% (in favour of the Liberal party) over this period. This study explores the question of how electoral characteristics relate to party preference, and whether their effects have changed over time.

Joining aggregate socio-demographics information from the Census with federal election voting data is problematic, as there in an inherent asynchronicity in the two events - a Census occurs every five years, whereas an election usually occurs every three years. The predominant approach in previous studies is to join voting outcomes to the nearest Census, without accounting for any temporal differences (see @DavisStimson98, @Stimson06, @Liao09 and @Stimson09). Furthermore, electoral boundaries change regularly, so spatial discrepencies also arise when matching electoral data. To obtain appropriate Census data for the 2004, 2007, 2010 and 2013 elections, we adopt impute electoral socio-demographics using spatio-temporal imputation that combines areal interpolation [@Goodchild1993] and linear time-interpolation. The process of collection and wrangling is detailed in @CHAPTER2 and imputation process is outlined in @CHAPTER3.

Previous work on modelling voting outcomes with aggregate socio-demographics find that relatively good models can be fit. @Forrest01 does this using multiple regression of the Liberal and Labor primary vote for polling booths in the Farrer electorate in 1998. @Stimson06, @Stimson09 and @Stimson12 conduct principal component analysis of polling booths in the 2001, 2004 and 2007 elections respectively, also finding that socio-demographic characteristics of polling booths are linked to their two-party preferred vote. @Stimson09 also model the polling booth swing vote (change in the two-party preferred vote) in the 2007 election, but had little success. Instead of analyzing a single election in isolation, this paper employs a consistent model framework across elections so that any temporal changes are observed.

@@

Joining aggregate socio-demographics information from the Census with federal election voting data is problematic, as there in an inherent asynchronicity in the two events - a Census occurs every five years, whereas an election usually occurs every three years. The predominant approach in previous studies is to join voting outcomes to the nearest Census, without accounting for any temporal differences (see @DavisStimson98, @Stimson06, @Liao09 and @Stimson09). Furthermore, electoral boundaries change regularly, so spatial discrepencies also arise when matching electoral data. To combat

@@

From our revivew of the literature, a regression framework using socio-demographics to understand Australian election outcomes has been applied on only two occasions. @Forrest01 use multiple regression to identify significant determinants of the primary vote for each party in the Farrer electorate during the 1998 election, by modelling each party's share of the primary vote as a separate regression, each using ordinary least squares. The adjusted R-squared values obtained from these regressions range from 0.158 to 0.785, suggesting that some parties (including the Labor and Liberal parties) are more predictable than others. They found that age, education and migrant populations were important in relation to support for both Labor and Liberal. @Stimson09 also apply a multiple regression framework to party preference, by modelling the swing vote in the 2007 election. The resultant model has relative poor fit (adjusted R-squared of 0.223), and as such no conclusions about socio-demographic effects were made. Both of these studies use a stepwise procedure to reduce the variable set, and had limited success in providing rich insights into socio-demographic preferences.

Applying a stepwise procedure to a data set with a low number of observations per predictor is problematic [@Steyerberg2009]. In our study, each election is modelled as a separate cross-section containing 63 socio-demographic predictors with only 150 observations. Therefore, instead of stepwise selection, we adopt an information-theoretic approach that uses Akaike weights to determine relative variable importance [@BurnhamAnderson2002], and use this as a criterion for variable selection. Akaike weights are often used in ecological studies (see @Johnson2004 and @Symonds2011).

>>>>

Australia has changed in many ways over the last two decades. Rising house prices, country-wide improvements in education, an ageing population, and a decline in religious affiliation, are just a few facets of the country's evolving socio-demographic characteristics[^1]. At the same time, political power has moved back and forth between the two major parties. In the 2007 and 2010 federal elections, the Australian Labor Party (Labor) was victorious, whereas the 2001, 2004, 2013 and 2016 elections were won by the Liberal National coalition (Liberal). The two-party preferred vote, a measure of support between these two parties, fluctuated between 47.3% and 53.5% (in favour of the Liberal party) over this period. This study explores how electoral characteristics relate to two-party preference, and whether their effects have changed over time. Information about electoral socio-demographics are from the Census, and vote counts from the federal election.

Joining these two data sources is problematic, as there in an inherent asynchronicity in the two events - a Census occurs every five years, whereas an election usually occurs every three years. The predominant approach in previous studies is to join voting outcomes to the nearest Census, without accounting for any temporal differences (see @DavisStimson98, @Stimson06, @Liao09 and @Stimson09). Furthermore, electoral boundaries change regularly, so spatial discrepancies also arise when matching electoral data. To obtain appropriate Census data for the 2004, 2007, 2010 and 2013 elections, we adopt impute electoral socio-demographics using spatio-temporal imputation that combines areal interpolation [@Goodchild1993] and linear time-interpolation. The process of collection and wrangling is detailed in @CHAPTER2 and imputation process is outlined in @CHAPTER3. All data and documentation relating to this procedure is available in the `eechidna` `R` package on `CRAN`.

Previous work on modelling voting outcomes with aggregate socio-demographics find that relatively good models can be fit. @Forrest01 does this using multiple regression of the Liberal and Labor primary vote for polling booths in the Farrer electorate in 1998. @Stimson06, @Stimson09 and @Stimson12 conduct principal component analysis of polling booths in the 2001, 2004 and 2007 elections respectively, also finding that socio-demographic characteristics of polling booths are linked to their two-party preferred vote. @Stimson09 also model the polling booth swing vote (change in the two-party preferred vote) in the 2007 election, but had little success. Instead of analyzing a single election in isolation, this paper employs a consistent model framework across elections so that any temporal changes are observed, where each of the six federal elections are modelled with a cross-sectional data set. This consists of the two-party preferred vote, which is the response variable, and socio-demographic variables which are explanatory variables that characterise each electorate. This process is discussed in @CHAPTER4 and involves standardizing variables across years, using principal components to group variables, and then selecting the most influential variables with an information-theoretic approach via Akaike weights to determine relative variable importance [@BurnhamAnderson2002]. Spatial error models are then fit to account for spatial dependence.

---------

@@

Joining aggregate socio-demographics information from the Census with federal election voting data is problematic, as there in an inherent asynchronicity in the two events - a Census occurs every five years, whereas an election usually occurs every three years. The predominant approach in previous studies is to join voting outcomes to the nearest Census, without accounting for any temporal differences (see @DavisStimson98, @Stimson06, @Liao09 and @Stimson09). Furthermore, electoral boundaries change regularly, so spatial discrepancies also arise when matching electoral data. To combat

@@

From our revive of the literature, a regression framework using socio-demographics to understand Australian election outcomes has been applied on only two occasions. @Forrest01 use multiple regression to identify significant determinants of the primary vote for each party in the Farrer electorate during the 1998 election, by modelling each party's share of the primary vote as a separate regression, each using ordinary least squares. The adjusted R-squared values obtained from these regressions range from 0.158 to 0.785, suggesting that some parties (including the Labor and Liberal parties) are more predictable than others. They found that age, education and migrant populations were important in relation to support for both Labor and Liberal. @Stimson09 also apply a multiple regression framework to party preference, by modelling the swing vote in the 2007 election. The resultant model has relative poor fit (adjusted R-squared of 0.223), and as such no conclusions about socio-demographic effects were made. Both of these studies use a step-wise procedure to reduce the variable set, and had limited success in providing rich insights into socio-demographic preferences.

Applying a step-wise procedure to a data set with a low number of observations per predictor is problematic [@Steyerberg2009]. In our study, each election is modelled as a separate cross-section containing 63 socio-demographic predictors with only 150 observations. Therefore, instead of step-wise selection, we adopt an information-theoretic approach that uses Akaike weights to determine relative variable importance [@BurnhamAnderson2002], and use this as a criterion for variable selection. Akaike weights are often used in ecological studies (see @Johnson2004 and @Symonds2011).
